---
title: "GGSEA with limma voom"
author: "Tycho Bismeijer"
date: "2017-12-06"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GGSEA with limma voom}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(42)
```


```{r generate-data, include = FALSE}
n_samples <- 10
n_genes <- 2500
spiked_geneset_size = 30
counts <- matrix(rnbinom(n_genes * n_samples, mu=1000, size=2), ncol=n_samples)
dimnames(counts) <- list(
    'gene'=paste0('g', 1:n_genes),
    'sample'=paste0('samples', 1:n_samples))
y <- data.frame(group=letters[sample.int(2, n_samples, replace=T)])
spiked_geneset = paste0('g', sample.int(n_genes, spiked_geneset_size))
counts[spiked_geneset, y == 'b'] = (counts[spiked_geneset, y == 'b'] + 
    rnbinom(spiked_geneset_size * sum(y == 'b'), mu=1000, size=10))
gene_sets <- sapply(paste0('gene_set', 2:10), function (i) {
    paste0('g', sample.int(n_genes, rnorm(1, 25, 10)))
}, simplify=F)
gene_sets[['gene_set1']] = spiked_geneset
```

We have generated a small random dataset with `r n_genes` genes and
`r n_samples` samples. The samples are divided into two groups: 'a' and 'b'.
We also generated `r length(gene_sets)` random gene sets. Gene set 1 is enriched
samples of group b, so should come out as significant.

```{r show data}
str(y)
str(counts)
```

First step in the gene set enrichment analysis with limma is to define the experiment design.
```{r design}
design <- model.matrix(~ group, y)
```

Second we use voom to convert to log2 count per million (CPM) and
calculate precision weights.
```{r voom}
dge <- edgeR::DGEList(counts)
dge <- edgeR::calcNormFactors(dge)
v <- limma::voom(dge, design)
```

Finally, we run GGSEA with `ggsea_limma` as the gene scoring function. 
```{r ggsea, message=F}
library(ggsea)
gse <- ggsea(v, design, gene_sets, ggsea_limma, nperm=1000)
```

The main results are contained in the `gse$table`, which is a list of the gene
set enrichments. As we have a single analysis (a versus b) in this example it is
list of length one. The list item is a `data.frame` with the statistics for
every gene set.

```{r str-table}
str(gse$table[[1]])
```

We can use dplyr to select the enriched gene sets.

```{r show-table}
library(dplyr)
gse$table[[1]] %>%
    filter(fdr < 0.25) %>%
    knitr::kable()
```
